Bayesian multi-proxy reconstruction of early Eocene latitudinal temperature gradients

### 2023-08-14 manuscript/main.qmd
review updates

To test whether our logistic regression model can adequately describe different latitudinal temperature gradients at various sample sizes, we generated four idealised gradients that emulate potential climatic states throughout Earth's geological history: extreme icehouse, icehouse, greenhouse, and extreme greenhouse [@Frakes1992]. We then randomly sampled (1,000 iterations) these gradients using increasing sample sizes (5, 10, and 20) and reconstructed the latitudinal temperature gradient using our model for each of these sample sizes and gradient types. Using the same idealised gradients, we also tested whether our model could accurately reconstruct latitudinal temperature gradients using the palaeogeographic distribution of Eocene samples (*n* = 34), providing an empirical, exemplary distribution that captures both limited sample size and skewed geographic origins of samples. To evaluate how well the model performed in reconstructing the idealised gradients from limited sampling, we calculated the coefficient of determination ($R^2$) for Bayesian regression models [@Gelman2019]. For every iteration from the posterior, we intercepted the modelled and the idealised gradient in intervals of 1&deg; latitude and calculated the $R^2$ based on these values. We report the median, and 95% credible intervals (CI) of the resulting $R^2$ values. Here and in all other instances, the 95% CI refer to the interval between the 2.5% point and the 97.5% point of the samples or sampled posterior distribution.

>>>>

To test whether our logistic regression model can adequately describe different latitudinal temperature gradients at various sample sizes, we used the empirical, modern gradient, representative of an icehouse climate, and generated three idealised gradients that emulate potential climatic states throughout Earth's geological history: extreme icehouse, icehouse (modern), greenhouse, and extreme greenhouse [@Frakes1992]. The idealised gradients serve to test whether our model setup is able to infer gradients that are strongly different from the modern from a varying amount of samples.

We created test data from these gradients as follows: To capture both limited sample size and skewed geographic origins of samples, we randomly sampled (200 iterations) the latitudes of the EECO data set (n = 34) used in this study, at sample sizes of 5, 10 and 20, and also used the full set of 34 latitudes. For each latitude, we inferred a location mean temperature from the gradients, adding random noise from a normal distribution with a standard deviation of 2.2, which is the residual standard deviation of modern sea surface temperature averages from the modelled modern gradient (see below). We then randomly allocated a sample size $s$ from the sample sizes present in the EECO data set to each location, and randomly generated $s$ temperature data from a normal distribution with mean equal to the previously generated location mean temperatures, and a standard deviation of 3.8. This standard deviation is the average standard deviation of the EECO geochemical proxy data, and with that we aim to simulate randomly distributed errors in the proxy data. This approach should give a fair assessment of how the model setup can be affected by the known unknowns (quantifiable errors), but we acknowledge that it cannot quantify the potential impact of unknown, systematic offsets that may bias all proxy data from one location in the same direction.

To evaluate how well the model performed in reconstructing the idealised gradients from limited sampling, we calculated the coefficient of determination ($R^2$) for Bayesian regression models [@Gelman2019]. For every iteration from the posterior, we intercepted the modelled and the idealised gradient in intervals of 1&deg; latitude and calculated the $R^2$ based on these values. We report the median, and 95% credible intervals (CI) of the resulting $R^2$ values. Here and in all other instances, the 95% CI refer to the interval between the 2.5% point and the 97.5% point of the samples or sampled posterior distribution.
